---
title: "Aizarani integration parameters"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(tidyverse)
source("workflow.R")
```
Exploring the different parameters used for integration of the Aizarani dataset with Marti's.  Sometimes when using the `FindTransferAnchors` function, I get the error:
```
Error in idx[i, ] <- res[[i]][[1]] : 
  number of items to replace is not a multiple of replacement length
```
And then R crashes. 

I 
This seems to depend on the choice of k.filter - some values work for some datasets but not others.

```{r include=FALSE}
save_file <- "aizarani_integration_test.rds"
if (!file.exists(save_file)) {
  print("loading data")
  aizarani <- import_aizarani()
  merged <- import_noMouse_merged()
  saveRDS(list(aizarani, merged), save_file)

  } else {
  print("importing data")
  aizarani_list <- readRDS(save_file)
  aizarani <- aizarani_list[[1]]
  merged <- aizarani_list[[2]]
  rm(aizarani_list)
}




```

```{r include=FALSE}
mapHepQuery <- function(query_name, k.filter=200) {
  print(glue::glue("working on {query_name}"))
  query <- merged[[query_name]]
  
  query <- NormalizeData(query, normalization.method = "LogNormalize", scale.factor = 10000)
  query <- FindVariableFeatures(query, selection.method = "vst", nfeatures = 2000)
  anchors <-  FindTransferAnchors(reference=aizarani, query=query,
                                   dims = 1:30, reference.reduction = "pca",
                                  k.filter=k.filter)
  query <-  MapQuery(anchorset = anchors, 
                       reference=aizarani,
                       query=query, 
                       refdata = list(paper.cluster.names="paper.cluster.names",
                                      paper.cluster="paper.clusters"), 
                       reference.reduction="pca",
                       reduction.model = "umap")
  p <- DimPlot(query, reduction="ref.umap", group.by="predicted.paper.cluster.names") + ggtitle(glue::glue("{query_name}, k.filter = {k.filter}"))
  return(list(p, query))
  }


#mapHepQuery(merged[['high']], k.filter=200)
# doesn't work


#k.filter <- 150
#mapHepQuery(merged[['high']], k.filter=k.filter)
# doesn't work

#purrr::map(merged, ~mapHepQuery(., k.filter=NA))


# tried values for all merged datasets:
#200 - didn't work
#150 - didn't work
#100 - didn't work
#300 - didn't work
#250 -- didn't work
#NA - works
#500 - doesn't work

```

How does changing the k.filter affect the results?

```{r include=FALSE}
refplot <- DimPlot(aizarani, reduction="umap", group.by="paper.cluster.names") + ggtitle("Aizarani reference") 
```


```{r include=FALSE}
hep.filt.200 <- mapHepQuery("hep", k.filter=200)
hep.filt.50 <- mapHepQuery("hep", k.filter=50)
hep.filt.NA <- mapHepQuery("hep", k.filter=NA)

```


```{r fig.height=7, fig.width=7}


cowplot::plot_grid(refplot + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10), 
                   hep.filt.200[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10), 
                   hep.filt.50[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10), 
                   hep.filt.NA[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10))
```

Different values of k.filter don't appear to change too much in this graph.

The 'low' sample only seems to run with a k.filter of 200, and the 'high', 'LK03C_REDHV' and 'LK03V_REDHC' samples only appear to run with a k.filter of NA.

What do these datasets look like?

```{r include=FALSE}
low.filt.200 <- mapHepQuery("low", k.filter=200)
high.filt.NA <- mapHepQuery("high", k.filter=NA)
LK03C_REDHV.filt.NA <- mapHepQuery("LK03C_REDHV", k.filter=NA)
LK03V_REDHC.filt.NA <- mapHepQuery("LK03V_REDHC", k.filter=NA)


```

```{r fig.height=7, fig.width=10}

cowplot::plot_grid(refplot + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10), 
                   hep.filt.200[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10),
                   low.filt.200[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10), 
                   high.filt.NA[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10),
                   LK03C_REDHV.filt.NA[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10),
                   LK03V_REDHC.filt.NA[[1]] + NoLegend() + xlim(-15, 15) + ylim(-12.5, 10)
                   )

```

Perhaps the integration failing with particular values of k.filter has something to do with how many cell clusters are represented?  The hep dataset, which has the most variablility in cell type, can be integrated with many different values.  For the other datasets, which have few cell types, only a k.filter NA or 200 works.

Also check the counts of each cell type:

```{r}
count_celltypes <- function(seurat_obj, name) {
  t <- tibble(
    seurat_obj[[2]][["predicted.paper.cluster.names"]]
  ) %>% 
    count(predicted.paper.cluster.names) %>% 
    setNames(c("cell type", name))
  
  return(t)
}

aizarani.counts <- tibble(aizarani[["paper.cluster.names"]]) %>% 
  count(paper.cluster.names) %>% 
  rename(aizarani = n)

lst <- list(
  "hep.filt.200" = hep.filt.200,
  'hep.filt.50' = hep.filt.50,
  'hep.filt.NA' = hep.filt.NA,
  'low.filt.200' = low.filt.200,
  'high.filt.NA' = high.filt.NA,
  'LK03C_REDHV.filt.NA' = LK03C_REDHV.filt.NA,
  'LK03V_REDHC.filt.NA' = LK03V_REDHC.filt.NA  
)

for (n in names(lst)) {
  counts <- count_celltypes(lst[[n]], n)
  aizarani.counts <- left_join(aizarani.counts, counts, by=c("paper.cluster.names"="cell type"))
}

aizarani.counts %>% 
  knitr::kable()
```


Try [integrating five references](https://github.com/joeb-liver/Single_Cell_Liver_Atlas/blob/main/Liver_Atlas_Integration.R), and then integrate each dataset with that?

## Only hepatocytes

Take the subset of the Aizarani dataset that are lablled as hepatocytes.  Check out the clusters that Aizarani found in this dataset, and try to find our own clusters.

```{r}
unique(aizarani$paper.cluster.names)

aizarani.hep <- subset(aizarani, paper.cluster.names=="Hepatocytes")

mapHepQuery <- function(reference, query_name, k.filter=200) {
  print(glue::glue("working on {query_name}"))
  query <- merged[[query_name]]
  
  query <- NormalizeData(query, normalization.method = "LogNormalize", scale.factor = 10000)
  query <- FindVariableFeatures(query, selection.method = "vst", nfeatures = 2000)
  anchors <-  FindTransferAnchors(reference=reference, query=query,
                                   dims = 1:30, reference.reduction = "pca",
                                  k.filter=k.filter)
  query <-  MapQuery(anchorset = anchors, 
                       reference=reference,
                       query=query, 
                       refdata = list(paper.cluster.names="paper.cluster.names",
                                      paper.cluster="paper.clusters"), 
                       reference.reduction="pca",
                       reduction.model = "umap")
  p <- DimPlot(query, reduction="ref.umap", group.by="predicted.paper.cluster.names") + ggtitle(glue::glue("{query_name}, k.filter = {k.filter}"))
  return(list(p, query))
}

mapHepQuery(aizarani.hep, "hep", k.filt=NA)

```

Integrating `aizarani.hep` with the hepatocytes sample:
```
Error: Please set k.weight to be smaller than the number of anchors (3).
```

Maybe it's better to integrate first, then take the subset.
