---
title: "R Notebook"
output:
  html_notebook:
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: hide
  html_document:
    keep_md: true
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height=5, dev = c('png', 'pdf'))
library(Seurat)
library(RColorBrewer)
library(tidyverse)
library(cowplot)

source("workflow.R")

cache_path <- "cache"
dir.create(cache_path)

legend_text_size <- 10

```
```{r loadIntegratedWithAizarani, include=FALSE}

cache_file <- file.path(cache_path, "integrated_with_aizarani.RDS")

if (!file.exists(cache_file)) {
  data <- integrated_with_aizarani()
  saveRDS(data, cache_file)
} else {
  data <- readRDS(cache_file)
}


reference <- data[['aizarani']]
queries <- data[names(data) != 'aizarani']
```

```{r loadIntegrated, include=FALSE}

cache_file <- file.path(cache_path, "integerated_together_no_mouse.RDS")

if (!file.exists(cache_file)) {

data.obj <- import_without_mouse()

## add column names for later
for (i in seq_along(names(data.obj))) {
  name <- names(data.obj)[i]
  type <- stringr::str_replace(name, "\\d+$", "")
  data.obj[[name]]$dataset.ind <- rep(i, length(colnames(data.obj[[name]])))
  data.obj[[name]]$dataset.name <-  rep(name, length(colnames(data.obj[[name]])))
  data.obj[[name]]$dataset.type <- rep(type, length(colnames(data.obj[[name]])))
  
  data.obj[[name]] <- NormalizeData(data.obj[[name]], normalization.method = "LogNormalize", scale.factor = 10000)
  data.obj[[name]] <- FindVariableFeatures(data.obj[[name]], selection.method = "vst", nfeatures = 2000)
}

# do integration
features <- SelectIntegrationFeatures(object.list = data.obj)
anchors <- FindIntegrationAnchors(object.list = data.obj, anchor.features = features, dims=1:30)
cells <- IntegrateData(anchorset = anchors, dims=1:30)
DefaultAssay(cells) <- "integrated"

# Run the standard workflow for visualization and clustering in the integrated data
cells <- ScaleData(cells, verbose = FALSE)
cells <- RunPCA(cells, npcs = 30, verbose = FALSE)
cells <- RunUMAP(cells, dims = 1:30, reduction.name = "UMAP")
cells <- FindNeighbors(cells, reduction = "pca", dims = 1:30)
cells <- FindClusters(cells, resolution = 0.5)

saveRDS(cells, cache_file)
} else {
  cells <- readRDS(cache_file)
}

```

```{r annotate, include=FALSE}
cells$cell.type <- case_when(
  cells$seurat_clusters == 0 ~ "Portal hepatoctyes (0)",
  cells$seurat_clusters == 1 ~ "Mid-zonal hepatocytes (1)",  
  cells$seurat_clusters == 2 ~ "Mid-zonal hepatocytes (2)",
  cells$seurat_clusters == 3 ~ "Central hepatocytes (3)",
  cells$seurat_clusters == 4 ~ "Portal hepatocytes (4)",
  cells$seurat_clusters == 5 ~ "Mid-zonal hepatocytes (5)",
  cells$seurat_clusters == 6 ~ "Endothelial cells (6)"
)

cells$cell.type.shorter <- case_when(
  cells$seurat_clusters == 0 ~ "Portal 0",
  cells$seurat_clusters == 1 ~ "Mid-zonal 1",  
  cells$seurat_clusters == 2 ~ "Mid-zonal 2",
  cells$seurat_clusters == 3 ~ "Central 3",
  cells$seurat_clusters == 4 ~ "Portal 4",
  cells$seurat_clusters == 5 ~ "Mid-zonal 5",
  cells$seurat_clusters == 6 ~ "Endothelial"
)

cells$cell.type.shortest <- case_when(
  cells$seurat_clusters == 0 ~ "Portal",
  cells$seurat_clusters == 1 ~ "Mid-zonal",  
  cells$seurat_clusters == 2 ~ "Mid-zonal",
  cells$seurat_clusters == 3 ~ "Central",
  cells$seurat_clusters == 4 ~ "Portal",
  cells$seurat_clusters == 5 ~ "Mid-zonal",
  cells$seurat_clusters == 6 ~ "Endothelial"
)

```


Here is the data where I integrated all the datasets against the Aizarani reference:

```{r AizIntegatedCellTypesAllDatasets, fig.width=7, fig.height=5}
cellTypes <- unique(reference$paper.cluster.names)
myColors <- brewer.pal(length(cellTypes),"Set3")
names(myColors) <- cellTypes
colScale <- scale_colour_manual(name = "grp",values = myColors)

refplot <- DimPlot(reference, reduction="umap", group.by="cell.types") + ggtitle("Aizarani") + colScale
queryplots.aizarani <- purrr::map(names(queries), 
                       ~DimPlot(queries[[.]], reduction="ref.umap", group.by="predicted.cell.type") + 
                         ggtitle(str_split(., "\\.")[[1]][1]) + 
                         xlim(-15, 15) + 
                         ylim(-12.5, 10) +
                         colScale +
                         NoLegend()
                       
                       )
# https://stackoverflow.com/questions/27130610/legend-on-bottom-two-rows-wrapped-in-ggplot2-in-r
legend.b <- cowplot::get_legend(refplot + 
                                theme(legend.position="bottom", 
                                      legend.box="vertical", text=element_text(size=legend_text_size), 
                                      legend.title = element_blank()) + 
                                guides(fill=guide_legend(nrow=5,byrow=TRUE))
                              )


aizarani.plots <- plot_grid(plotlist=c(list(refplot + NoLegend()), queryplots.aizarani))
plot_grid(aizarani.plots, legend.b, ncol=1, rel_heights = c(1, 0.1))
```


How many cells are there of each type?

```{r}
tibble(
  sample = names(queries)
) %>% 
  rowwise() %>% 
  mutate(data = list(queries[[sample]]@meta.data)) %>% 
  unnest(data) %>% 
  count(sample, predicted.cell.type) %>% 
  pivot_wider(names_from = "predicted.cell.type", values_from = n)
  
```
In the above table I combined the two replicates for each sample type, because I integrated them with the reference together.  To get them separate counts for each replicate, I use the cell names in the original data.

```{r}
integrated_together <- cells@meta.data
integrated_together$cell_id <- rownames(integrated_together)

get_metdata_with_cell_ids <- function(sample) {
  data <- queries[[sample]]@meta.data
  data$cell_id <- rownames(data)
  return(data)
}


integrated_aizarani <- tibble(
  sample = names(queries)
) %>% 
  rowwise() %>% 
  mutate(data = list(get_metdata_with_cell_ids(sample))) %>% 
  unnest(data) 

integrated_aizarani %>% 
  left_join(dplyr::select(integrated_together, cell_id, dataset.name), by="cell_id") %>% 
  count(dataset.name, predicted.cell.type) %>% 
  pivot_wider(names_from = "predicted.cell.type", values_from = n)

```




And here is the data where I integrated all the datasets against each other:


```{r fig.height=10, fig.width=10}
Idents(cells) <- cells$seurat_clusters
p1 <- DimPlot(cells) + ggtitle("Seurat Clusters")
Idents(cells) <- cells$cell.type.shorter
p2 <- DimPlot(cells) + ggtitle("Named clusters")
Idents(cells) <- cells$cell.type.shortest
p3 <- DimPlot(cells) + ggtitle("Combined clusters")
cowplot::plot_grid(p1, p2, p3)
```

I assume that all the cells that aren't hepatocytes in the first figure are the cluster down the bottom in the second figure (cluster 6).  Here I check how many cells from the first figure are in each of the clusters from the second figure.

In the table below, the columns 0 through 6 are the "Seurat clusters" from the second figure, and the predicted.cell.type column is the cell type from the first figure.

```{r}

integrated_together %>% 
  left_join(dplyr::select(integrated_aizarani, cell_id, predicted.cell.type), by="cell_id") %>% 
  count(dataset.name, seurat_clusters, predicted.cell.type) %>% 
  pivot_wider(names_from = seurat_clusters, values_from = n)

```


For comparison, here are the overall number of cells in clusters 0-6:

```{r}
integrated_together %>% 
  count(dataset.name, seurat_clusters)  %>% 
  pivot_wider(names_from = seurat_clusters, values_from = n)
```


Try to plot this - on the UMAP plot from the second figure, color cells by their predicted cell type from the Aizarani integration

```{r}
cell_types <- tibble(
  cell_id = colnames(cells)
) %>% 
  left_join(select(integrated_aizarani,predicted.cell.type, cell_id), by="cell_id")



cells$aizarani.cell.type <- cell_types$predicted.cell.type


Idents(cells) <- cells$aizarani.cell.type
p4 <- DimPlot(cells) + ggtitle("Predicted cell type")

p4

```



