---
title: "R Notebook"
output:
  html_document:
    keep_md: true
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    code_folding: hide
---

Try running `learn_graph` from `Monocle3` on Marti's datasets after integration to find a trajectory.

```{r importData, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height=5, dev = c('pdf', 'png'))

library(Seurat)
library(dplyr)
library(patchwork)
source("workflow.R")


data.obj <- import_without_mouse()
save.file <- glue::glue("../out/Seurat/integrated/mouse-{params$include_mouse}")
dir.create(dirname(save.file), recursive=TRUE)


## add column names for later
for (i in seq_along(names(data.obj))) {
  name <- names(data.obj)[i]
  type <- stringr::str_replace(name, "\\d+$", "")
  data.obj[[name]]$dataset.ind <- rep(i, length(colnames(data.obj[[name]])))
  data.obj[[name]]$dataset.name <-  rep(name, length(colnames(data.obj[[name]])))
  data.obj[[name]]$dataset.type <- rep(type, length(colnames(data.obj[[name]])))
  
  data.obj[[name]] <- NormalizeData(data.obj[[name]], normalization.method = "LogNormalize", scale.factor = 10000)
  data.obj[[name]] <- FindVariableFeatures(data.obj[[name]], selection.method = "vst", nfeatures = 2000)
}



# standard workflow
features <- SelectIntegrationFeatures(object.list = data.obj)
anchors <- FindIntegrationAnchors(object.list = data.obj, anchor.features = features, dims=1:30)
cells <- IntegrateData(anchorset = anchors, dims=1:30)
DefaultAssay(cells) <- "integrated"

# Run the standard workflow for visualization and clustering
cells <- ScaleData(cells, verbose = FALSE)
cells <- RunPCA(cells, npcs = 30, verbose = FALSE)

cells <- RunUMAP(cells, dims = 1:30, reduction.name = "UMAP")

cells <- FindNeighbors(cells, reduction = "pca", dims = 1:30)
cells <- FindClusters(cells, resolution = 0.5)
```


I first integrate the datasets together using Seurat.  The Seurat clusters in the integrated dataset look like this:

```{r plotSeuratClust, fig.height=8, fig.width=6}
plotlist <- DimPlot(cells, group.by=c("ident", "dataset.type", "dataset.name"), combine=FALSE)
plotlist <- purrr::map(plotlist, ~. + ggplot2::theme(legend.position = "bottom"))
cowplot::plot_grid(plotlist=plotlist)
```

```{r makeMonocleClust include=FALSE}

library(monocle3)
library(SeuratWrappers)

cells.monocle <- as.cell_data_set(cells)
cells.monocle <- cluster_cells(cells.monocle)

p1 <- plot_cells(cells.monocle, show_trajectory_graph = FALSE)
p2 <- plot_cells(cells.monocle, color_cells_by = "partition", show_trajectory_graph = FALSE)


```





These are the monocle clusters

```{r monocleClusters}
wrap_plots(p1, p2)
```

```{r monoclePseudotime}
cells.monocle <- learn_graph(cells.monocle)
plot_cells(cells.monocle, label_groups_by_cluster = FALSE, label_leaves = TRUE, label_branch_points = TRUE)
```

I'm not sure how to interpret this trajectory - it appears to be somewhat circular.  We have to manually define where the trajectory begins to calculate pseudotime, but I'm not sure how to do that.




